
---
title: "Flexible Modeling and Competing Risks"
author: "Faida Clemence NIYONSHIMIYE"
date: "`r Sys.Date()`"
output: html_document

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```





```{r}
library(survival)
library(rstpm2)

data(lung, package = "survival")

str(lung)
head(lung)
```
```{r}
lung$status2 <- ifelse(lung$status == 2, 1, 0)

table(lung$status, lung$status2)
```

```{r}
#Fit the Cox and Royston-Parmar models
cox_fit <- coxph(Surv(time, status2) ~ sex + age, data = lung)
summary(cox_fit)

rp_fit <- stpm2(Surv(time, status2) ~ sex + age, data = lung, df = 4)
summary(rp_fit)
```

```{r}
#Compare coefficient estimates
cox_coef <- coef(cox_fit)
rp_coef  <- coef(rp_fit)[c("sex", "age")]

cox_coef
rp_coef
```
```{r}
#Cox cumulative baseline hazard
cox_bh <- basehaz(cox_fit, centered = FALSE)

head(cox_bh)
```
```{r}
#Royston-Parmar hazard estimates over time
tgrid <- seq(min(lung$time), max(lung$time), length.out = 200)

newdat <- data.frame(
  time = tgrid,
  sex  = 1,
  age  = mean(lung$age, na.rm = TRUE)
)

rp_haz <- predict(rp_fit, newdata = newdat, type = "hazard")
rp_df <- data.frame(time = tgrid, hazard = as.numeric(rp_haz))
head(rp_df)
```

```{r}
# -----------------------------
# 1. Cox cumulative baseline hazard plot
# -----------------------------
cox_bh <- basehaz(cox_fit, centered = FALSE)

plot(cox_bh$time, cox_bh$hazard,
     type = "s",
     col = "blue",
     lwd = 2,
     xlab = "Time",
     ylab = "Cumulative baseline hazard",
     main = "Cox Model: Cumulative Baseline Hazard")

# -----------------------------
# 2. Royston-Parmar hazard plot
# -----------------------------

plot(tgrid, rp_haz,
     type = "l",
     col = "red",
     lwd = 2,
     xlab = "Time",
     ylab = "Hazard",
     main = "Royston-Parmar Model: Hazard Function")
```
```{r}
#on the same Graph
# -----------------------------
# 1. Cox cumulative baseline hazard plot
# -----------------------------
cox_bh <- basehaz(cox_fit, centered = FALSE)

plot(cox_bh$time, cox_bh$hazard,
     type = "s",
     col = "blue",
     lwd = 2,
     xlab = "Time",
     ylab = "Cumulative baseline hazard",
     main = "Cox Model: Cumulative Baseline Hazard")

# -----------------------------
# 2. Royston-Parmar hazard plot
# -----------------------------
lines(tgrid, rp_haz,
      col = "red",
      lwd = 2)

legend("topleft",
       legend = c("Cox cumulative baseline hazard", "Royston-Parmar hazard"),
       col = c("blue", "red"),
       lwd = 2,
       lty = c(1, 1),
       bty = "n")
```

## Interpretation
The Cox hazard appears as a step function because the Cox model is semi-parametric, so the baseline hazard changes only at the observed event times.

The Royston--Parmar hazard is smooth because it models the baseline hazard flexibly using splines.

Overall, the Royston--Parmar curve is easier to interpret visually, while the Cox curve reflects the observed data more directly.


